Embedding Intuitionistic into Classical Logic

### 2023-03-10 LPAR/LPAR-EIICL.tex
Lemma 4.4

In this section we fix notation and recall the semantics for classical and intuitionistic logic. 
Furthermore, we recapitulate some results from~\cite{otten2005clausal}.

@@

For predicate logic we assume an ambient signature $\Sigma$, i.e., a set of function and predicate symbols with their respective arities, and a set of variables. A term is either a variable or, given an $n$-ary function symbol $f$ and terms $t_1, \ldots, t_n$, $f(t_1, \ldots, t_n)$. A formula is either $\bot$, or, given a $n$-ary predicate symbol $P$ and terms $t_1, \ldots, t_n$, $P(t_1, \ldots, t_n)$, or, given formulas $\varphi$ and $\psi$ and $\circ\in\{\wedge, \vee,\to\}$, $\varphi\circ\psi$, or, given a formula $\varphi$ and a variable $x$, $\exists x\varphi$ or $\forall x\varphi$. Free and bound variables are defined as usual and we assume that they comprise distinct sets, such that bound variables are not affected by substitutions of free variables. Generally letters at the beginning of the alphabet will denote free variables and letters at the end of the alphabet bound variables.

@@

\begin{definition}
	Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the \emph{domain} of $\mathcal{M}$, and an \emph{interpretation} $I$ that assigns
	\begin{itemize}
		\item to each $n$-ary function symbol $f\in\Sigma$ a $n$-ary function $f^I: M^n\to M$.
		\item to each $n$-ary predicate symbol $R\in\Sigma$ a $n$-ary predicate $R^I\subseteq M^n$.
	\end{itemize}
	A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
	m, &\text{if $b=a$,}\\
	v(b), &\text{otherwise.}
	\end{cases}$$
	Then terms are interpreted as follows:
	\begin{itemize}
		\item $a^{I, v} = v(a)$ for each free variable $a$.
		\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$ for each $n$-ary function symbol $f\in\Sigma$.
	\end{itemize}
	We define a model relation between pairs $\mathcal M, v$ and formulas $\varphi$ as follows
	\begin{itemize}
		\item $\mathcal M, v\not\models\bot$
		\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$, $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
		\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$
		\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/a]\models\varphi[a/x]$.
		\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/a]\models\varphi[a/x]$.\todo{Hier und im Rest vom paper bin nicht so glücklich über diese Substitution von durch a..}
	\end{itemize}
	 Here $a$ is a free variable that does not occur in $\varphi$. $\varphi$ is satisfiable if $\mathcal M, v\models\varphi$ for some $\mathcal M, v$. It is valid if $\mathcal M, v\models\varphi$ for every $\mathcal M, v$. We denote the set of valid formulas with \emph{CQC} (Classical Quantified Calculus).
\end{definition}

@@

An important step in the embedding will be the elimination of quantifiers via Herbrandization.
In this process we introduce fresh variables and add additional function symbols to the signature.
A fresh variable is any variable that does not occur in any considered formula.
Whenever we add a function symbol we implicitly extend the signature by a fresh symbol.

\begin{definition}\label{definition:skolemization}
	For formulas $\varphi$ we define the Skolemization $\varphi^S_Z$ and Herbrandization $\varphi^H_Z$ with respect to a set of free variables $Z$ by simultaneous induction:
	\begin{itemize}
		\item $A^S_Z = A^H_Z = A$ for each atomic $A$.
		\item $(\varphi\circ\psi)^X_Z = \varphi^X_Z\circ\psi^X_Z$ for $\circ\in\{\wedge, \vee\}$, $X\in\{S, H\}$.
		\item $(\varphi\to\psi)^S_Z = \varphi^H_Z\to \psi^S_Z$ and $(\varphi\to\psi)^H_Z = \varphi^S_Z\to\psi^H_Z$.
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi[a/x]^S_{Z\cup\{a\}}[x/a])$ where $a$ is a new free variable.\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.\\$(\exists x\varphi)^H_Z = \exists x(\varphi[a/x]^H_{Z\cup\{a\}}[x/a])$
 where $a$ is a new free variable.
	\end{itemize}
	Let $\varphi^S = (\exists x_1\dots\exists x_n \varphi[x_1/a_1\dots x_n/a_n])^S_\emptyset$ and $\varphi^H = (\forall x_1\dots\forall x_n \varphi[x_1/a_1\dots x_n/a_n])^H_\emptyset$ where $a_1,\dots,a_n$ are the free variables occurring in $\varphi$.
\end{definition}

@@

Finally we recall a result from~\cite{otten2005clausal}, on which we build in this paper.

\begin{lemma}\label{lemma:propositional-normal-form}
	For every propositional formula $\varphi$ there exists an atom $P$ as well as sets of clauses $\mathcal R, \mathcal X$, with $\mathcal R$ containing  \emph{flat clauses} of the form
	$$\bigwedge_iA_i\to\bigvee_jB_j$$
	and $\mathcal X$ containing \emph{implication clauses} of the form
	$$(A\to B)\to C,$$
	such that $\varphi$ is intuitionistically equivalid to
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$where $A_i, B_i, A, B, C$ are atomic. The size of $\mathcal R$ and $\mathcal X$ is linear in the size of $\varphi$. In general these formulae will contain additional propositional variables.
\end{lemma}

\begin{lemma}\label{lemma:first-order-normal-form}
	For every predicate formula $\varphi$ there exists a nullary predicate symbol $P$ as well as sets of clauses $\mathcal R,\mathcal X, \mathcal Q$, with $\mathcal R$ containing \emph{flat clauses} of the form
	$$\forall \vec x\left(\bigwedge_i A_i\to \bigvee_jB_j\right),$$
	$\mathcal X$ containing \emph{implication clauses} of the form
	$$\forall \vec x\left((A\to B)\to C\right)$$
	and $\mathcal Q$ containing \emph{quantification clauses} of the form
	$$\forall\vec x\left((\forall y A)\to B\right),$$
	such that $\varphi$ is intuitionistically equivalid to
	$$\left(\bigwedge\mathcal R\wedge\bigwedge \mathcal X\wedge\bigwedge\mathcal Q\right)\to P$$where $A_i, B_i, A, B, C$ are atomic. This size of $\mathcal R, \mathcal X, \mathcal Q$ is linear in the size of $\varphi$. In general these formulae will contain new function and predicate symbols.
\end{lemma}

Since we could not find a proof for the first-order case, we provide one in the appendix~\ref{proof:first-order-normal-form}.

@@

We shall now give an overview of our main results and highlight the key arguments. 
We start with the propositional case, as the important insights are the same as in the first-order case, while it is technically simpler and less cluttered.

The first step is to express the Kripke semantics of intuitionistic logic in classical first-order logic.
That is, for each $\varphi$ we give a $\varphi^\#$ that is classically valid iff $\varphi$ is intuitionistically valid, establishing a dual to the double negation translation.

Then, the key to reducing the first-order logic encoding to a propositional logic encoding, is a small model property that supports the reduction of counter-models.
In particular, given $$\varphi = \left(\bigwedge\mathcal R\wedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form} we establish:

@@

This allows us to embed intuitionistic propositional logic into classical propositional logic by replacing the quantification over all possible worlds by an explicit enumeration:

@@

The details can be found in Section~\ref{sec:ipc-to-cpc}.
Building on the same idea, we can also give a quantified boolean formula (QBF) of polynomial size, where the branches of the tree-model can be enumerated implicitly using (alternating) quantifiers. We obtain the following result:

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a QBF $\varphi^Q$ \todo{Ich würde die Formeln $\varphi^Q$ hier und in Section 6 aus Symmetriegründen auch $\varphi^\#$ nennen.} with $\mathcal |\varphi^Q|\in\mathcal O(|\mathcal R|\cdot|\mathcal X| + |\mathcal X|^3)$ such that $\varphi$ is intuitionistically invalid if and only if $\varphi^Q$ is a satisfiable QBF.
\end{lemma}

Again, the details can be found in Section~\ref{sec:qbf}. 
Next we attempt to transport these results to the first order case. However, there are formulas for which only counter-examples with an infinite Kripke frame exist. Therefore, there is no hope of completely eliminating terms representing worlds from the final embedding. This also means that our language necessarily becomes more complex. Still, a number of reductions are possible.

@@

\begin{definition}
	Let $\varphi$ be a propositional formula.\todo{genau genommen müsstest du noch sagen, dass $\nu$ eine freie Variable ist. Im letzten Fall der Definition benützen wir dann aber w gleichzeitig als freie und gebundene Variable}
 Define $\varphi^{\nu}$ inductively:
	\vspace*{-.1cm}
	\begin{itemize}
		\item $\bot^\nu = \bot$.
		\item $A^{\nu} = A(\nu)$ for every propositional variable $A$.
		\item $(\varphi\circ\psi)^\nu = \varphi^\nu\circ\psi^\nu$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^\nu = \forall w(\nu\preceq w\to\varphi^{w}\to\psi^{w})$ where $w$ is some new bound variable.
	\end{itemize}
	Let $K(\varphi)$ encode the theory of Kripke structures, i.e.
	\vspace*{-.1cm}
	\[K(\varphi) = \text{\normalfont PartialOrder}(\preceq)\wedge\forall u\forall w(u\preceq w\to \text{\normalfont Persistent}(u, w))\]
	\vspace*{-.4cm}
	with e.g.
	\vspace*{-.2cm}
	\begin{align*}
		\text{\normalfont PartialOrder}(\preceq) =&\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{\normalfont Persistent}(u, w)=&\:\bigwedge \{A(u)\to A(w)\:|\: \text{ $A$ occurs in $\varphi$}\}
	\end{align*}
	Then define
	$\varphi^{C} = K(\varphi)\to \varphi^{b}$
	where $b$ is a new free variable.
\end{definition}

@@

This is somewhat unsatisfying as we end up in a much more complex logic. However using the normal form from Lemma~\ref{lemma:propositional-normal-form}, i.e. assuming that $\varphi$ is of the form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$ for some flat clauses $\mathcal R$ and implicational clauses $\mathcal X$, further simplification is possible:
\begin{itemize}
	\item For flat clauses $(\bigwedge A_i\to\bigvee B_j)^\nu$ is
	$\forall w\left(\nu\preceq w\to\bigwedge_i A_i(w)\to\bigvee_jB_j(w)\right).$
	\item For implication clauses $((A\to B)\to C)^\nu$ is
	$$\forall w(\nu\preceq w\to(\forall k(w\preceq k\to A(k)\to B(k)))\to C(w))$$
\end{itemize}

and after applying Herbrandization we end up with a formula $$\varphi^{CH} = K(\varphi)\to\left(\bigwedge\mathcal R^{b S}\wedge \bigwedge\mathcal X^{b S}\right)\to P(s)$$ where $s$ is a new constant symbol, each formula in $\mathcal R^{b S}$ is of the form
$$\forall w\left(s\preceq w\to\bigwedge_iA_i(w)\to\bigvee_jB_j(w)\right)$$
and each formula $\psi\in\mathcal X^{b S}$ is of the form
$$\forall w(s\preceq w\to(w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w)))\to C(w))$$
with $f_\psi$ being a new function symbol.
We now argue that if a counter-model for $\varphi^{CH}$ exists then a small counter-model exists. This is ultimately what enables our translations.

\begin{definition}
For $\psi\in\mathcal X$ let $[\psi]$ denote the corresponding clause in $\mathcal X^{b S}$. Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$.
We say $(A\to B)\to C\in\mathcal X$ is \emph{fulfilled} at $m\in M$ iff $A^I(m)\to B^I(m)$ is false or $C^I(m)$ is true. For $\psi\in\mathcal X$ define $$g_\psi : M\to M, m\mapsto\begin{cases}
		m,&\text{ if $\psi$ is fulfilled at $m$,}\\
		f^I_{\lbrack\psi\rbrack}(m),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal X$.
		\item $s^{I_T} = \epsilon$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $[\psi_1]\dots[\psi_n]$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}\left(\psi_1\dots \psi_n\right) = P^I\left(g_{\psi_n}\left(\dots\left(g_{\psi_1}\left(s^I\right)\right)\dots\right)\right).$$
	\end{itemize}
\end{definition}
\begin{lemma}~\label{thm:prop-countermodel-reduction}
	Let $\mathcal M = (M, I)$ be a counter-model to $\mathcal \varphi^{CH}$.
	\begin{enumerate}
		\item $\psi$ is fulfilled at $f_{[\psi]}^I(m)$ for all $m\in M, \psi\in\mathcal X$.
		\item If $\psi$ is fulfilled at some $m\in M$ then $\psi$ is fulfilled at all $n\succeq^I m$.
	\end{enumerate}
\end{lemma}

\begin{proof}
	1. If $C^I(m)$ is true then we are done due to persistency. Otherwise $C^I(m)$ is false.
	Because $(A^I(f_{[\psi]}^I(m))\to B^I(f_{[\psi]}^I(m)))\to C^I(u)$ holds, then $A^I(f_{[\psi]}^I(u))\to B^I(f_{[\psi]}^I(u))$ must be false.

	2. Let $n\succeq^I m$.
	If $C^I(n)$ is true, we are done.
	Otherwise $C^I(n)$ is false.
	Due to persistency $C^I(m)$ is also false.
	Because $\psi$ is fulfilled at $m$, $A^I(m)\to B^I(m)$ must be false, i.e. $A^I(m)$ is true and $B^I(m)$ is false. Then $A^I(n)$ and $A^I(f^I_{[\psi]}(n))$ are also true due to persistency.
	Because $(A^I(f^I_{[\psi]}(n))\to B^I(f^I_{[\psi]}(n)))\to C^I(n)$ holds, we now get that  $B^I(f^I_{[\psi]}(n))$ must be false.
	But then due to persistency so is $B^I(n)$ and we have that $\psi$ is fulfilled at $n$.
\end{proof}

Essentially what this tells us is that for every $m$ at which $\psi$ is fulfilled, in particular for each $m$ in the image of $f_{[\psi]}$, we may add the assumption that $f_{[\psi]}(n) = n$ for all $n\succeq m$ without changing the validity of $\psi$. With this lemma in hand showing that $(M_T, I_T)$ is a counter-model to $\varphi^{CH}$ is a straightforward check of definitions.

@@

\section{IPC to CPC}
\label{sec:ipc-to-cpc}

@@

It is known that a polynomial time translation between QBF and IPC must exist, since both problems are PSPACE complete~\cite{garey1979computers,statman1979intuitionistic}.
\todo{Wollen wir noch etwas sagen wie ``However the previous PSPACE result does not support the transfer of countermodels that we develop in this section''?}
For lack of space, we do not give the semantics of QBF in this paper and instead refer the reader to standard sources e.g.~\cite{series/faia/2009-185}.
The translation from the last section will serve as starting point for this section.

Suppose we are given a formula $\varphi$ in normal form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form}. We now given an intuition on how to obtain a QBF of polynomial size that expresses that $\varphi$ has an intuitionistic counter-model: 
Instead of expressing validity at every node of $\mathcal M_T$ --- which was done in the CPC translation --- we express that on each path in $\mathcal M_T$ the nodes satisfy all conditions for $\mathcal M_T$ being a counter-model. 
The universally quantified variables $X_\psi$ in the next definition handle the branching, i.e. express which path is considered.

>>>>

In this section we fix notation and recall the semantics for classical and intuitionistic logic. Furthermore we recapitulate some results from~\cite{otten2005clausal}.

@@

For predicate logic we assume an ambient signature $\Sigma$, i.e., a set of function and predicate symbols with their respective arities, and a set of variables. A term is either a variable or, given an $n$-ary function symbol $f$ and terms $t_1, \ldots, t_n$, $f(t_1, \ldots, t_n)$. A formula is either $\bot$, or, given a $n$-ary predicate symbol $P$ and terms $t_1, \ldots, t_n$, $P(t_1, \ldots, t_n)$, or, given formulas $\varphi$ and $\psi$ and $\circ\in\{\wedge, \vee,\to\}$, $\varphi\circ\psi$, or, given a formula $\varphi$ and a variable $x$, $\exists x\varphi$ or $\forall x\varphi$. Free and bound variables are defined as usual and there is a Barendregt's convention in place, i.e. we implicitly assume that they comprise distinct sets, such that bound variables are not affected by substitutions of free variables.

@@

\begin{definition}
	Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the \emph{domain} of $\mathcal{M}$, and an \emph{interpretation} $I$ that assigns
	\begin{itemize}
		\item to each $n$-ary function symbol $f\in\Sigma$ a $n$-ary function $f^I: M^n\to M$.
		\item to each $n$-ary predicate symbol $R\in\Sigma$ a $n$-ary predicate $R^I\subseteq M^n$.
	\end{itemize}
	A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
	m, &\text{if $b=a$,}\\
	v(b), &\text{otherwise.}
	\end{cases}$$
	Then terms are interpreted as follows:
	\begin{itemize}
		\item $a^{I, v} = v(a)$ for each free variable $a$.
		\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$ for each $n$-ary function symbol $f\in\Sigma$.
	\end{itemize}
	We define a model relation between pairs $\mathcal M, v$ and formulas $\varphi$ as follows
	\begin{itemize}
		\item $\mathcal M, v\not\models\bot$
		\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$, $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
		\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$
		\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/a]\models\varphi[a/x]$.
		\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/a]\models\varphi[a/x]$.
	\end{itemize}
	 Here $a$ is a free variable that does not occur in $\varphi$. $\varphi$ is satisfiable if $\mathcal M, v\models\varphi$ for some $\mathcal M, v$. It is valid if $\mathcal M, v\models\varphi$ for every $\mathcal M, v$. We denote the set of valid formulas with \emph{CQC} (Classical Quantified Calculus).
\end{definition}

@@

An important step in the embedding will be the elimination of quantifiers via Herbrandization.
In this process we introduce fresh variables and add additional function symbols to the signature.
A fresh variable is any variable that does not occur in any considered formula.
Whenever we add a function symbol we implicitly extend the signature by a not previously contained symbol.

\begin{definition}\label{definition:skolemization}
	For formulas $\varphi$ we define the Skolemization $\varphi^S_Z$ and Herbrandization $\varphi^H_Z$ with respect to a set of free variables $Z$ by simultaneous induction:
	\begin{itemize}
		\item $A^S_Z = A^H_Z = A$ for each atomic $A$.
		\item $(\varphi\circ\psi)^X_Z = \varphi^X_Z\circ\psi^X_Z$ for $\circ\in\{\wedge, \vee\}$, $X\in\{S, H\}$.
		\item $(\varphi\to\psi)^S_Z = \varphi^H_Z\to \psi^S_Z$ and $(\varphi\to\psi)^H_Z = \varphi^S_Z\to\psi^H_Z$.
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi^S_{Z\cup\{x\}})$.\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.\\$(\exists x\varphi)^H_Z = \exists x(\varphi^H_{Z\cup\{x\}})$.
	\end{itemize}
	Let $\varphi^S = (\exists x_1\dots\exists x_n \varphi[x_1/a_1\dots x_n/a_n])^S_\emptyset$ and $\varphi^H = (\forall x_1\dots\forall x_n \varphi[x_1/a_1\dots x_n/a_n])^H_\emptyset$ where $a_1,\dots,a_n$ are the free variables occurring in $\varphi$.
\end{definition}

@@

Finally we note the results from~\cite{otten2005clausal} which are of importance to us.

\begin{lemma}\label{lemma:propositional-normal-form}
	For every propositional formula $\varphi$ there exists an atom $P$ as well as sets of clauses $\mathcal R, \mathcal X$, with $\mathcal R$ containing  \emph{flat clauses} of the form
	$$\bigwedge_iA_i\to\bigvee_jB_j$$
	and $\mathcal X$ containing \emph{implication clauses} of the form
	$$(A\to B)\to C,$$
	such that $\varphi$ is intuitionistically equivalid to
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$where $A_i, B_i, A, B, C$ are atomic. The size of $\mathcal R$ and $\mathcal X$ is linear in the size of $\varphi$. In general they will contain additional propositional variables.
\end{lemma}

\begin{lemma}\label{lemma:first-order-normal-form}
	For every predicate formula $\varphi$ there exists a nullary predicate symbol $P$ as well as sets of clauses $\mathcal R,\mathcal X, \mathcal Q$, with $\mathcal R$ containing \emph{flat clauses} of the form
	$$\forall \vec x\left(\bigwedge_i A_i\to \bigvee_jB_j\right),$$
	$\mathcal X$ containing \emph{implication clauses} of the form
	$$\forall \vec x\left((A\to B)\to C\right)$$
	and $\mathcal Q$ containing \emph{quantification clauses} of the form
	$$\forall\vec x\left((\forall y A)\to B\right),$$
	such that $\varphi$ is intuitionistically equivalid to
	$$\left(\bigwedge\mathcal R\wedge\bigwedge \mathcal X\wedge\bigwedge\mathcal Q\right)\to P$$where $A_i, B_i, A, B, C$ are atomic. This size of $\mathcal R, \mathcal X, \mathcal Q$ is linear in the size of $\varphi$. In general this formula will contain new function and predicate symbols.
\end{lemma}

Since we could not find a proof of this, one is given in the appendix~\ref{proof:first-order-normal-form}.

@@

We shall now give an overview of our main results and highlight the key arguments. We start with the propositional case. The important insights are the same as in the first-order case, while it is technically simpler and less cluttered.

The idea is to use the Kripke semantics for intuitionistic logic, expressed in classical logic, to perform an embedding of formulas. That is, for each $\varphi$ we give a $\varphi^\#$ that is classically valid iff $\varphi$ is intuitionistically valid, establishing a dual to the double negation translation.

The key property that allows us to reduce our encoding to propositional logic, which a priori targets first order logic, is a reduction of counter-models. In particular, given $$\varphi = \left(\bigwedge\mathcal R\wedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form} we establish:

@@

This allows us to embed intuitionistic propositional logic into classical propositional logic by replacing quantification over worlds by their enumeration.

@@

The exact details can be found in Theorem~\ref{thm:prop-translation}. Instead of expressing correctness at each world of the model, we can express that the worlds along some path in the tree are correct. The branching of paths can be expressed through quantification in a quantified boolean formula (QBF) which gives us the following:

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a QBF $\varphi^Q$ with $\mathcal |\varphi^Q|\in\mathcal O(|\mathcal R|\cdot|\mathcal X| + |\mathcal X|^3)$ such that $\varphi$ is intuitionistically invalid if and only if $\varphi^Q$ is a satisfiable QBF.
\end{lemma}

Again, the details can be found in Section~\ref{sec:qbf}. Next we attempt to transport these results to the first order case. However, there are formulas for which only counter-examples with an infinite Kripke frame exist. Therefore, there is no hope of completely eliminating terms representing worlds from the final embedding. This also means that our language necessarily becomes more complex. Still, a number of reductions is possible.

@@

\begin{definition}
	Let $\varphi$ be a propositional formula. Define $\varphi^{\nu}$ inductively:
	\vspace*{-.1cm}
	\begin{itemize}
		\item $\bot^\nu = \bot$.
		\item $A^{\nu} = A(\nu)$ for every propositional variable $A$.
		\item $(\varphi\circ\psi)^\nu = \varphi^\nu\circ\psi^\nu$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^\nu = \forall w(\nu\preceq w\to\varphi^{w}\to\psi^{w})$ where $w$ is some new bound variable.
	\end{itemize}
	Let $K(\varphi)$ encode the theory of Kripke structures, i.e.
	\vspace*{-.1cm}
	\[K(\varphi) = \text{\normalfont PartialOrder}(\preceq)\wedge\forall u\forall w(u\preceq w\to \text{\normalfont Persistent}(u, w))\]
	\vspace*{-.4cm}
	with e.g.
	\vspace*{-.2cm}
	\begin{align*}
		\text{\normalfont PartialOrder}(\preceq) =&\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{\normalfont Persistent}(u, w)=&\:\bigwedge \{A(u)\to A(w)\:|\: \text{ $A$ occurs in $\varphi$}\}
	\end{align*}
	Then define
	$\varphi^{C} = K(\varphi)\to \varphi^{b}$
	where $b$ is a new free variable.
\end{definition}

@@

This is somewhat unsatisfying as we end up in a much more complex logic. However using the normal form from Lemma~\ref{lemma:propositional-normal-form}, i.e. assuming that $\varphi$ is of the form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$ for some flat clauses $\mathcal R$ and implicational clauses $\mathcal X$, further simplification is possible:
\begin{itemize}
	\item For flat clauses $(\bigwedge A_i\to\bigvee B_j)^\nu$ is
	$$\forall w\left(\nu\preceq w\to\bigwedge_i A_i(w)\to\bigvee_jB_j(w)\right).$$
	\item For implication clauses $((A\to B)\to C)^\nu$ is
	$$\forall w(\nu\preceq w\to(\forall k(w\preceq k\to A(k)\to B(k)))\to C(w))$$
\end{itemize}

and after applying Herbrandization we end up with a formula $$\varphi^{CH} = K(\varphi)\to\left(\bigwedge\mathcal R^{b S}\wedge \bigwedge\mathcal X^{b S}\right)\to P(s)$$ where $s$ is a new constant symbol, each formula in $\mathcal R^{b S}$ is of the form
$$\forall w\left(s\preceq w\to\bigwedge_iA_i(w)\to\bigvee_jB_j(w)\right)$$
and each formula $\psi\in\mathcal X^{b S}$ is of the form
$$\forall w(s\preceq w\to(w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w)))\to C(w))$$
with $f_\psi$ being a new function symbol.

\begin{definition}
For $\psi\in\mathcal X$ let $[\psi]$ denote the corresponding clause in $\mathcal X^{b S}$. Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$ such that $m\preceq f_\psi^I(m)$ for each $m\in M$, $\psi\in \mathcal X$.
We say $(A\to B)\to C\in\mathcal X$ is \emph{fulfilled} at $m\in M$ iff $A^I(m)\to B^I(m)$ is false or $C^I(m)$ is true. For $\psi\in\mathcal X$ define $$g_\psi : M\to M, m\mapsto\begin{cases}
		m,&\text{ if $\psi$ is fulfilled at $m$ or $m\not\preceq f^I_{[\psi]}(m)$,}\\
		f^I_{\lbrack\psi\rbrack}(m),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal X$.
		\item $s^{I_T} = \epsilon$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $[\psi_1]\dots[\psi_n]$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}\left(\psi_1\dots \psi_n\right) = P^I\left(g_{\psi_n}\left(\dots\left(g_{\psi_1}\left(s^I\right)\right)\dots\right)\right).$$
	\end{itemize}
\end{definition}

\begin{lemma}
	Suppose $(M, T)$ is a counter-model to $\varphi^{CH}$, then so is $(M_T, I_T)$.
\end{lemma}
\begin{proof}
	It follows directly from the definitions that $M_T, I_T\not\models P(s)$ and $M_T, I_T\models K(\varphi)$. It is easy to show that $M_T, I_T\models \psi$ for $\psi\in\mathcal R^{bS}$ from persistency. Let $\psi\in\mathcal X^{bS}$, we show $M_T, I_T[\psi_1\dots\psi_n/w] \models s\preceq w\to (w\preceq f_\psi(w)\to A(f_\psi(w)\to B(f_\psi(w)))\to C(w))$ for all $\psi_1\dots\psi_n\in M_T$. Let $\psi_1\dots\psi_n\in M_T$ and $m = g_{\psi_n}(\dots(g_{\psi_1}(s^I))$. If $M, I[m/w]\models C(w)$ we are done. Otherwise $M, I[m/w]\not\models s\preceq w$ or $M, I[m/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w)) \to B(f_\psi(w))$. The former is impossible since $g_\psi$ is increasing for all $\psi$. For the latter to hold we must have $M, I[m/w]\models w\preceq f_\psi(w)$, $M, I[m/w]\models A(f_\psi(w))$ and $M, I[m/w]\not\models B(f_\psi(w))$. If $\psi$ is fulfilled at $m$ we have $g_\psi(m) = m$, otherwise $g_\psi(m) = f_{[\psi]}^I(m)$. In any case $M, I'[\psi_1\dots\psi_n/w]\models A(f_\psi(w))$ and $M_T, I_T[\psi_1\dots\psi_n/w]\not\models B(f_\psi(w))$. It follows that $M_T, I_T[\psi_1\dots\psi_n/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w))$ and therefore $M_T, I_T\models \psi$.
\end{proof}

@@

\section{IPC to CPC}

@@

It is known that a polynomial time translation between QBF and IPC must exist, since both problems are PSPACE complete~\cite{garey1979computers,statman1979intuitionistic}.
The translation from the last section will serve as starting point for developing the translation in this section.
For lack of space, we do not define the semantics of QBF in this paper and refer the reader to standard sources e.g.~\cite{series/faia/2009-185}.

Suppose we are given a formula $\varphi$ in normal form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form}. We will now formulate a QBF that expresses that $\varphi$ has an intuitionistic counter-model. The idea how to get a polynomially-sized formula is: Instead of expressing validity at every node of $\mathcal M_T$ --- which was done in the CPC translation --- we express that on each path in $\mathcal M_T$ the nodes satisfy all conditions for $\mathcal M_T$ being a counter-model. The universally quantified variables $X_\psi$ in the next definition handle the branching, i.e. express which path is considered.
